Synthesis of Rules for Network Intrusion Detectors

### 2019-10-24 paper/paper.tex
pass

Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.
\begin{wrapfigure}[16]{r}{0.5\textwidth}
  \centering
  \vspace{-5ex}
  \scalebox{0.7}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-rules-protocol}Number of relevant options per protocol.}
\end{wrapfigure}
The purpose of rule option \CodeIn{msg}, for instance, is to print on
the output an informative message indicating what kind of intrusion
has been observed. Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. Note that the sum of
the number of protocol options is higher than \totoptions\ as some
options can be used by multiple protocols. Table~\ref{table:rules}
shows a sample of the options of Suricata. Column ``\#'' shows the
option number, column ``Name'' shows the name of the option, column
``\#~Params'' shows the number of additional parameters associated
with a given option, and column ``Description'' presents a short
description of the option purpose. \Fix{...explain content...}

\begin{table}[h]
  \caption{\label{table:rules}Example options.}  
  \centering
  \begin{tabular}{p{0.35cm}ll}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & matches on bytes of the payload\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Rules and Packets}

@@

\Fix{atualize isto para o valor correto, por favor. $\rightarrow$}

\begin{figure}[h!]
  \vspace{-3ex}
  \lstinputlisting[language=C,numbers=none,frame=none,keywords={flags,threshold}]{synflood.suricata.synth}
%  \caption{Suricata rule for SYN Flood Attacks.}
  %  \label{fig:synflood-example-synt}
  \vspace{-3ex}  
\end{figure}

This set is based on an analysis of the packets from the positive
traffic. Consequently, a rule that uses this set of options would
capture the positive traffic by construction. Unfortunately, such rule
would be overspecified, \ie{}, the rule would unlikely capture the
same attacks as the traffic data would not be the same and \tname{}
includes in the set all possible options that can be inferred from the
data. To address that, \tname{} minimizes the rule with the goal of
finding the minimal set of rules that only captures the positive
traffic, hence mitigating the overfitting problem. \tname{}
potentially finds several minimal rules. The list below shows the
options for the five shortest rules it produces for this attack.

@@

\subsection{\Fix{...some other with contents...}}

@@

\vspace{1ex}
\noindent\textbf{Overview.}~ The input of the technique is the
positive traffic whereas the output is a list of minimized rules
capturing only the positive, malicious,
traffic. Figure~\ref{fig:overview} shows the workflow of \tname{} as a
pipeline of two components, which appear numbered in the figure.
\tname{} synthesizes rules in two steps. First, it produces a
potentially overspecified rule that captures the malicious traffic
provided on input. In this step, \tname\ finds options by
\emph{reverse engineering} the positive traffic. It extracts tokens
from the malicious message and uses them to define the initial
rule. We say that this rule overfits the problem as it satisfies the
success criterion of only capturing the positive traffic, however it
is unlikely that it will be able to generalize to other attacks (say,
attacks of the same kind, but with different data). The reason is that
many options added to the rule at the first step are not important for
discriminating the attack. In the second step, \tname{} uses negative
examples to \emph{minimize} the rule, discarding unnecessary
options. Note that any subset of the original set of options captures
the positive traffic. The minimization step uses a population-based
gradient descent-like search to generate minimal solutions.

@@

@@

The goal of the first step is to extract options from the positive
traffic. To infer options, \tname{} parses the protocol message, looking for
fields in the message associated with options. For example, fields
\Fix{...and...}  can be used to determine the option
\Fix{...}. Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content} options. Figure~\ref{fig:distribution-contents} shows
the histogram of number of these options per rule (up to 10) for the
Suricata public rule database \Fix{cite}. Rules without \CodeIn{content}
options is an exception and half of the rules contain at least two
\CodeIn{content} options.
\begin{wrapfigure}[14]{r}{0.56\textwidth}
  \centering
  \vspace{-3ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
To find the values associated with these options, \tname{} splits the
payload of the malicious message in tokens using natural language
delimiters, such as $\backslash$t, $\backslash$n, $\backslash$r, and
spaces. The rationale for this decision is that we observed, by
inspecting existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. \Fix{describe extraction of options not in the
  message. give examples! search only makes sense if this exists!}

>>>>

Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.
\begin{wrapfigure}[15]{r}{0.5\textwidth}
  \centering
  \vspace{-5ex}
  \scalebox{0.7}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-1ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
\end{wrapfigure}
\Comment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, column ``\#~Params'' shows the number of
additional parameters associated with a given option, and column
``Description'' presents a short description of the option purpose. We
found that the option \CodeIn{content} is particularly prevalent in
rules from public databases (see
Figure~\ref{fig:distribution-contents}). This option takes a sequence
of bytes as argument (\eg{}, \CodeIn{content:''SELECT ''}) and checks
if that sequence is present in the payload of the message. Detailed
description of the Suricata rules can be found
elsewhere~\cite{suri-rule-format}.

\begin{table}[h]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{p{0.35cm}ll}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Rules and Packets}
\label{sec:rules-and-packets}

@@

\Fix{atualize isto para o valor correto, por favor. $\rightarrow$}
\begin{figure}[h]
  \vspace{-3ex}
  \lstinputlisting[language=C,numbers=none,frame=none,keywords={flags,threshold}]{synflood.suricata.synth}
%  \caption{Suricata rule for SYN Flood Attacks.}
  %  \label{fig:synflood-example-synt}
  \vspace{-3ex}  
\end{figure}

This set is based on an analysis of the packets from the positive
traffic. Consequently, a rule that uses this set of options would
capture the positive traffic by construction. Unfortunately, such rule
would be overspecified. It would unlikely capture different
manifestations of the same attack. Informally, the rule includes ``too
many'' characteristics of the malicious traffic used to define it, but
new attacks may not use the same data. To address that, \tname{}
minimizes the rule with the goal of finding the minimal set of rules
that only captures the positive traffic, hence mitigating the
overfitting problem. \tname{} potentially finds several minimal
rules. The list below shows the options for the five shortest rules it
produces for this attack.

@@

\subsection{\Fix{...some other with contents...}}
\label{sec:content-example}

@@

\vspace{1ex}
\noindent\textbf{Overview.}~ The input of the technique is the
positive traffic whereas the output is a list of minimized rules
capturing only the positive, malicious,
traffic. Figure~\ref{fig:overview} shows the workflow of \tname{} as a
pipeline of two components, which appear numbered in the figure.
\tname{} synthesizes rules in two steps. First, it produces a
potentially overspecified rule that captures the malicious traffic
provided on input. In this step, \tname\ finds options by
\emph{reverse engineering} the positive traffic. It extracts tokens
from the malicious message and uses them to define the initial
rule. We say that this rule overfits the problem as it satisfies the
success criterion of only capturing the positive traffic, however it
is unlikely that it will be able to generalize to other attacks,
\ie{}, attacks of the same kind, but with different data. The reason
is that many options added to the rule are unimportant for
discriminating the attack. In the second step, \tname{} uses negative
examples to \emph{minimize} the rule, discarding unnecessary
options. Note that any subset of the original set of options captures
the positive traffic. The minimization step uses a population-based
gradient descent-like search to generate minimal solutions.

@@

\begin{wrapfigure}[14]{r}{0.56\textwidth}
  \centering
  \vspace{-8ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
The goal of the first step is to extract options from the positive
traffic. For that, \tname{} parses the protocol message, looking for
fields in the message associated with options (see
Section~\ref{sec:rules-and-packets}). For example, fields
\Fix{...and...}  can be used to determine the option
\Fix{zzz}. Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content}
options. Figure~\ref{fig:distribution-contents} shows the histogram of
number of these options per rule (up to 10) for the Suricata public
rule database \Fix{cite}. Rules without \CodeIn{content} options is an
exception and half of the rules contain at least two \CodeIn{content}
options.  To find the values associated with these options, \tname{}
splits the payload of the malicious message in tokens using natural
language delimiters, such as $\backslash$t, $\backslash$n,
$\backslash$r, and spaces. The rationale for this decision is that we
observed, by inspecting existing rules, that \CodeIn{content} options
typically refer to sequence of characters in the payload of messages
separated by these delimiters. This tokenization step can generate
dozens of options. For the example from
Section~\ref{sec:content-example}, \tname{} produces a total of
\Fix{xx} content options.

@@
